### 2013-10-26 src2/thesis.tex
update

Modern CPUs are parallel in a number of different ways. They discover instruction
level parallelism (ILP), they have multiple processor cores and vector units for 
Single Instruction Multiple Data (SIMD)  parallelism. The exact configuration, number
of cores and number of vector units, varies over available processor models. 
To obtain maximum performance, with a specific target processor in mind, the programmer
can use low-level intrinsics in order to get SIMD parallelism and a threading library
for parallelism over the available cores. One problem with this is that the resulting program 
would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
system approaches this problem by providing a high level programming interface~\citet{ARBB2011}.
This interface is based on a set of parallel operations on arrays, such as map, folds 
and scans. ArBB was implemented as an embedded language in C++, but a low-level C interface to 
its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
target the available CPU or accelerator (such as the Intel Larrabee~\citet{Larrabee}, 
knights Ferry, Xeon Phi). In chapter~\ref{chap:ArBB}, there are papers describing work 
towards embedding Intel ArBB in Haskell.

@@

In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
for general purpose workloads. A dialect of C, called CUDA C, was released specifically 
to simplify implementation of general purpose algorithms on GPUs.

@@

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
local shared memory. The maximum block size is currently 1024 threads. Since 
the number of threads in a block can be higher than the warp size, a barrier 
synchronisation mechanism exists to ensure that threads within a block 
can communicate safely via the shared memory.

@@

This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
element wise sums into the result array, {\tt r}. The {\tt threadIdx} variable identifies 
a thread within a block. There is also a {\tt blockIdx} variable that identifies 
the block a thread and {\tt blockDim.x} specifies the number of threads per block.

In CUDA there is a notion of {\em host} and {\em device}, the device is the GPU and 
its memory (Global memory, usually a few Gigabytes). The host refers to the system housing 
the GPU. The host runs a control program that launches kernels on the GPU. First memory 
needs to be allocated in the device to hold the inputs and outputs. Then data is copied from 
the host systems memory into device memory.

@@

\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
many 400 to 600 clock cycles if the data is not cached. The programmer should 
use shared memory as much as possible to avoid redundant loading from global memory.
Shared memory could also be used to coalesce reads from global memory. Shared memory is 
a limited resource in the MP, so if a kernel uses very much shared memory it means 
fewer such blocks can be active on the MP.

@@

\item {\bf Avoid different execution paths within a block:} Any branching instructions that 
causes the execution path diverge within a warp will affect performance. When the execution 
paths within a warp diverges the computation is serialised.

\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
when synchronising within any conditionally executed code; every thread must reach this barrier.
Failing to ensure this will likely lead to the kernel producing incorrect answers. Note that 
this is a major limitation to the composability of CUDA functions. If it is safe or 
not to call some function at a certain place in your code, can only be decided by inspecting 
that functions implementation. 
\end{itemize}

>>>>

Modern CPUs are parallel in a number of different ways. They discover instruction
level parallelism (ILP), they have multiple processor cores and vector units for 
Single Instruction Multiple Data (SIMD)  parallelism. The exact configuration, number
of cores and number of vector units, varies over available processor models. 
To obtain maximum performance, with a specific target processor in mind, the programmer
can use low-level intrinsics in order to get SIMD parallelism and a threading library
for parallelism over the available cores. One problem with this is that the resulting program 
would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
system approaches this problem by providing a high level programming interface~\citet{ARBB2011}.
This interface is based on a set of parallel operations on arrays, such as map, folds 
and scans. ArBB was implemented as an embedded language in C++, but a low-level C interface to 
its capabilities was also provided. ArBB programs are JIT-ed (Just-In-Time compiled) to 
target the available CPU or accelerator (such as the Intel Larrabee~\citet{Larrabee}, 
knights Ferry, Xeon Phi). In chapter~\ref{chap:ArBB}, there are papers describing work 
towards embedding Intel ArBB in Haskell.

@@

In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
for general purpose programming. A dialect of C, called CUDA C, was released specifically 
to simplify implementation of general purpose algorithms on GPUs.

@@

Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
also divided into {\em Blocks}; the threads within a block can communicate using the 
shared memory. The maximum block size is currently 1024 threads. Since 
the number of threads in a block can be higher than the warp size, a barrier 
synchronisation mechanism exists to ensure that threads within a block 
can communicate safely via the shared memory.

@@

This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
element wise sums into the result array, {\tt r}. The {\tt threadIdx} variable identifies 
a thread within a block. There is also a {\tt blockIdx} variable that identifies 
the block a thread belongs to and {\tt blockDim.x} specifies the number of threads per block.

In CUDA there is a notion of {\em host} and {\em device}, the device is the GPU and 
its memory (Global memory, usually a few Gigabytes). The host refers to the system housing 
the GPU. The host runs a control program that launches kernels on the GPU. First memory 
needs to be allocated in the device to hold the inputs and outputs. Then data is copied from 
the host's system memory into device memory.

@@

\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
many 400 to 600 clock cycles if the data is not cached. The programmer should 
use shared memory as much as possible to avoid redundant loading from global memory.
Shared memory can also be used to coalesce reads from global memory. Shared memory is 
a limited resource in the MP, so if a kernel uses very much shared memory it means 
fewer such blocks can be active on the MP.

@@

\item {\bf Avoid different execution paths within a block:} Any branching instructions that 
causes the execution paths to diverge within a warp will affect performance. When the execution 
paths within a warp diverges the computation is serialised.

\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
when synchronising within any conditionally executed code; every thread must reach this barrier.
Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
this is a major limitation to the composability of CUDA functions. If it is safe or 
not to call some function at a certain place in your code, can only be decided by inspecting 
that functions implementation. 
\end{itemize}
